A molecular dynamics study of the solvation of CO\textsubscript{2}, water, and some organic compounds in the ionic liquids [emim][B(CN)\textsubscript{4}] and [emim][NTf\textsubscript{2}]

### 2018-02-21 ionic_liquids.tex
computational methods

\section{Computational details}
\label{sec:sim_details}

The gradual insertion/deletion of charged solutes involves scaling both the van der Waals (vdW) and the electrostatic intermolecular interaction. In order to avoid numerical instabilities associated to the vdW scaling, a standard approach is to employ a soft-core potential with a nonlinear coupling, and in this paper we use the modified Lennard-Jones potential due to Beutler~\textit{et al.}~\cite{Beutler_1994}. The electrostatic interactions, on the other hand, can be handled simply by a linear scaling~\cite{Naden_2015}. Furthermore, the contributions to $\Delta G$ originating at the different types of coupling can be taken into account separately, as follows
\begin{equation}
\begin{split}
\Delta G_{\text{sim}} =& \; \Delta G_{\,\text{vdW}} (\lambda_{\,\text{vdW}} = 0\rightarrow  1, \lambda_{\, \text{Coul}} = 0) \\
+ \; &\Delta G_{\,\text{Coul}} (\lambda_{\, \text{vdW}} = 1,\lambda_{\,\text{Coul}} = 0 \rightarrow 1),
\end{split}
\end{equation}
with $\lambda_{\,\text{vdW}}$ and $\lambda_{\,\text{Coul}}$ being the coupling parameters of the soft-core and electrostatic potentials, respectively. Note that in the process of coupling the vdW interactions, the Coulombic interactions are turned-off, while we set $\lambda_{\,\text{vdW}}$ = 1 when computing $\Delta G_{\,\text{Coul}}$.  This results in the ``end-point'' states given by ($\lambda_{\,\text{vdW}}$=0,$\lambda_{\,\text{Coul}}$=0) and ($\lambda_{\,\text{vdW}}$=1,$\lambda_{\,\text{Coul}}$=1), as expected. The complete sets of the coupling parameters we employ here are $\lambda_{\text{vdW}}$ = $\{0.0$, $0.05$, $0.1$, $0.2$, $0.25$, $0.3$, $0.35$, $0.4$, $0.45$, $0.5$, $0.6$, $0.7$, $0.8$, $0.9$, $1.0\}$ and $\lambda_{\text{Coul}}$ = $\{0.0$, $0.1$, $0.4$, $0.7$, $0.1\}$.

In this work, we perform independent simulation at each state of $\lambda_{\text{Coul}}$, while the expanded ensemble (EE) method~\cite{Lyubartsev_1992} is applied for sampling the states associated to the $\lambda_{\,\text{vdW}}$ set. In short, an EE is constructed as a weighted sum of subensembles, each assuming a different value of the variable chosen to explore the free energy landscape, which in our case is $\lambda_{\,\text{vdW}}$. This way, the sampling of the EE involves both sampling each subensamble with fixed variables (N,P,T,$\lambda_{\,\text{vdW}}$) and transitions in $\lambda_{\,\text{vdW}}$. Thus, the remarkable feature of the EE method is the possibility of exploring the whole alchemical route in a single simulation. In this work, we perform a sequential sampling of the EE \cite{Chodera_2011_2}, which consists in sampling each subensamble via NPT MD for a determined number of steps $N^{\text{EE}}$ followed with an attemp to change $\lambda_{\,\text{vdW}}$ using MC, and so on. In addition, the set of (arbitrary) weights $\eta$ of the subensambles enters in the acceptance criteria of MC, so it is possible to tune $\eta$ in order to enhance the sampling. If differences $\eta_i$ - $\eta_j$ were closed to $\Delta G_{ji}$, we expect an uniform sampling in $\lambda_{\,\text{vdW}}$~\cite{Lyubartsev_1992}. Of course, this can be achieved only iteratively. As demonstrated in Ref.~\citenum{Chodera_2011_2}, smaller values of $N^{EE}$ generates more decorrelated configurations of the physical variables. Therefore, we employ $N^{EE}$ = 5. This value, together with the chosen $\lambda_{\, \text{vdW}}$ set resulted to be satisfactory, as up to three preliminar simulations of the EE were enough to determine the weights required for a uniform sampling of $\lambda_{\, \text{vdW}}$. Those simulations were about 10 ns. The time at each $\lambda_{\text{Coul}}$ was also 10 ns, while the final simulations of the EE were in the range of 20-40 ns, depending on the system. Finally, MBAR \cite{Shirts_2008} was applied form the sample of each $\lambda$ to estimate the free energy.

We implemented in LAMMPS~\cite{Plimpton1995} the expanded ensemble methodology, as well as the codes regarding the \textit{soft-core} and DSF potential with linear coupling. All of this is available at https://github.com/atoms-ufrj/USER-ALCHEMICAL. The electrostatic interactions were calculated using the damped shifted force (DSF) method~\cite{Fennell2006} with $\alpha$ = 0.2 {\AA}$^{-1}$, with the exception of hexane, for which we considered full Coulombic interactions. For the systems modelled as rigid molecules, we employ the NPT method of Kamberaj~\textit{et al.}~\cite{Kamberaj_2005} using $\Delta t$ = 1 fs. For flexible systems, we employ the scheme due to Shinoda~\textit{et al}.~\cite{Shinoda2004}. Initial configurations were build using the PLAYMOL software available at https://github.com/atoms-ufrj/playmol.

Among the force fields described in Sec.~\ref{sec:force_field_il}, only in Ref.~\citenum{Koller_2012} one can find the atomic coordinates of a minimum energy configuration obtained in an \textit{ab-initio} simulation. For the other models, we employ the tool \textit{build} of PLAYMOL, which enables to specify the geometric data of each model: bond and angle lengths and dihedral angles, from which PLAYMOL automatically determines the Cartesian coordinates. We first carried out simulations of the isolated ion pairs in order to obtain the minimum energy configurations with respect to the corresponding (classical) model for each case. Then, we obtain configurations of minimum energy with respect to the corresponding classical model. For this, we perform a series of MD simulations of each cation-anion pair. These simulations were performed in a cubic box of length L = 50 {\AA}. Initially, a short DM NVT with a time-step of 0.1 fs, followed by an equilibration using $\Delta t$ = 0.5 fs and, finally, we minimize the energy using 1 fs in a process of ``quenching'', in which the temperature is gradually reduce until reaching a value close to 0 K, for which we employ the Nos\'{e}-Hoover thermostat~\cite{Martyna1992}.

@@

\subsection{Simulation strategies of ionic liquids}
\label{sec:prel_results}

In this section, we present results of preliminary simulations, which aimed at setting up a (more) computationally efficient protocol for the ionic liquids studied here. As alternatives to standard approaches, we assess the possibility of modeling the ions as rigid bodies and employing the DSF method~\cite{Fennell2006} for calculating the electrostatic interactions. For that, we compare the simulated properties of density and radial distribution functions to that obtained when considering the ions as flexible molecules and the Ewald method for the Coulombic interactions.

>>>>

\section{Computational methods}
\label{sec:sim_details}

With the aim of setting up a (more) computationally efficient protocol for the ILs, in this work we assess the possibility of modeling the ions as rigid bodies and employing the damped shifted force (DSF) method~\cite{Fennell2006} for computing the electrostatic interactions. As an alternative to the Ewald schemes, McCann~\textit{et al.}~\cite{McCann_2013} tested the DSF method for various ILs modeled in the OPLS-AA framework. They concluded that using $\alpha$ = 0.2 {\AA}$^{-1}$ and a cutoff distance ($r_c$) of 15 {\AA} provides the best energetic agreement with Ewald. In this work, the comparisons are made in terms of the simulated densities and radial distribution functions for which, as will be shown in the next section, $\alpha$ = 0.2 {\AA}$^{-1}$ and $r_c$ = 12 {\AA} resulted in good agreement for the ILs studied here. Those values were also employed for the charged solutes, with the exception of hexane, for which we considered undamped Coulombic interactions.

When it was possible to treat the ions as rigid bodies, the NPT algorithm of Kamberaj~\textit{et al.}~\cite{Kamberaj_2005} was employed to carry out the MD simulations. Otherwise, the scheme due to Shinoda~\textit{et al}.~\cite{Shinoda2004} was used along with the application of SHAKE~\cite{Ryckaert1977} to constrain the C-H bonds lengths. For both integrators, we employed $\Delta t$ = 1 fs.

The gradual insertion/deletion of charged solutes involves scaling both the van der Waals (vdW) and the electrostatic intermolecular interactions. In order to avoid numerical instabilities associated to the vdW scaling, a standard approach is to employ a soft-core potential with a nonlinear coupling, and in this paper we use the modified Lennard-Jones potential due to Beutler~\textit{et al.}~\cite{Beutler_1994}. The electrostatic interactions, on the other hand, can be handled simply by a linear scaling~\cite{Naden_2015} of the potential, which in this work corresponds to the DSF method~\cite{Fennell2006}.

Furthermore, the contributions to $\Delta G$ originating at the different types of coupling can be taken into account separately, as follows
\begin{equation}
\begin{split}
\Delta G_{\text{sim}} =& \; \Delta G_{\,\text{vdW}} (\lambda_{\,\text{vdW}} = 0\rightarrow  1, \lambda_{\, \text{Coul}} = 0) \\
+ \; &\Delta G_{\,\text{Coul}} (\lambda_{\, \text{vdW}} = 1,\lambda_{\,\text{Coul}} = 0 \rightarrow 1),
\end{split}
\end{equation}
with $\lambda_{\,\text{vdW}}$ and $\lambda_{\,\text{Coul}}$ being the coupling parameters of the soft-core and electrostatic potentials, respectively. Note that in the process of coupling the vdW interactions, the Coulombic interactions are turned-off, while we set $\lambda_{\,\text{vdW}}$ = 1 when computing $\Delta G_{\,\text{Coul}}$.  This results in the ``end-point'' states given by ($\lambda_{\,\text{vdW}}$=0,$\lambda_{\,\text{Coul}}$=0) and ($\lambda_{\,\text{vdW}}$=1,$\lambda_{\,\text{Coul}}$=1), as expected. The complete sets of the coupling parameters we employ here are $\lambda_{\text{vdW}}$ = $\{0.0$, $0.05$, $0.1$, $0.2$, $0.25$, $0.3$, $0.35$, $0.4$, $0.45$, $0.5$, $0.6$, $0.7$, $0.8$, $0.9$, $1.0\}$ and $\lambda_{\text{Coul}}$ = $\{0.0$, $0.1$, $0.4$, $0.7$, $0.1\}$.

In this work, we performed independent simulation at each state of $\lambda_{\text{Coul}}$, while the expanded ensemble (EE) method~\cite{Lyubartsev_1992} was applied for sampling the states associated to the $\lambda_{\,\text{vdW}}$ set. In short, an EE is constructed as a weighted sum of subensembles, each assuming a different value of the variable chosen to explore the free energy landscape, which in our case is $\lambda_{\,\text{vdW}}$. This way, the sampling of the EE involves both sampling each subensamble with fixed variables (N,P,T,$\lambda_{\,\text{vdW}}$) and transitions in $\lambda_{\,\text{vdW}}$. Thus, the remarkable feature of the EE method is the possibility of exploring the whole alchemical route in a single simulation. In this work, we perform a sequential sampling of the EE \cite{Chodera_2011_2}, which consists in sampling each subensamble via NPT MD for a determined number of steps $N^{\text{EE}}$ followed with an attemp to change $\lambda_{\,\text{vdW}}$ using MC, and so on. Independent sampling of $\lambda$~\cite{Chodera_2011_2}. In addition, the set of (arbitrary) weights $\eta$ of the subensambles enters in the acceptance criteria of MC, so it is possible to tune $\eta$ in order to enhance the sampling. If differences $\eta_i$ - $\eta_j$ were closed to $\Delta G_{ji}$, we expect an uniform sampling in $\lambda_{\,\text{vdW}}$~\cite{Lyubartsev_1992}. Of course, this can be achieved only iteratively. The EE procedure, together with the \textit{soft-core} potential and the DSF method with linear coupling were all implemented in the LAMMPS software package~\cite{Plimpton1995}.

As demonstrated in Ref.~\citenum{Chodera_2011_2}, smaller values of $N^{EE}$ generate more decorrelated samples in each subensemble. In this work, we employ $N^{EE}$ = 5 which, together with the specified $\lambda_{\, \text{vdW}}$ set, resulted to be satisfactory as up to 3 preliminary runs of the EE were enough to determine the weights $\eta$ that yielded, in a final simulation, a ``sufficient'' sampling of each subensamble. By ``sufficient'', we mean that the sampling provides an acceptable error in $\Delta G$. The simulation times for the preliminary and final EE runs were 10 and 40 ns, respectively, and each $\lambda_{\text{Coul}}$ state was simulated for 10 ns. Finally, the MBAR method~\cite{Shirts_2008}, as implemented by Klimovich~\textit{et al.}~\cite{Klimovich_2015}, was applied to estimate $\Delta G_{\text{vdW}}$ and $\Delta G_{\text{Coul}}$ .

We end this section with a brief description of the procedure for generating the initial configurations. Only Ref.~\citenum{Liu_2014} provides the atomic coordinates of the cation-anion pair corresponding to a minimum energy configuration obtained in an \textit{ab-initio} simulation. For the other models, we employed the tool \textit{build} of PLAYMOL~\cite{playmol}, a software integrated with PACKMOL~\cite{Mart_nez_2009}, which enables to specify the following geometric data: bond and angle lengths and dihedral angles, from which PLAYMOL automatically determines the Cartesian coordinates. We first carried out NVT MD simulations of the isolated ion pairs in order to obtain a minimum energy conformation for the corresponding (classical) model of each pair. These simulations were performed using the the Nos\'{e}-Hoover thermostat~\cite{Martyna1992} and a cubic box of length L = 50 {\AA} with non-periodic boundary conditions. To relax the initial configurations, a short run with a small time step of 0.1 fs was performed, which was followed by an equilibration using $\Delta t$ = 0.5 fs. Finally, we carried out a ``quenching'' simulation with $\Delta t$ = 1 fs, in which the temperature is gradually reduced until reaching a value close to 0 K. The resulting coordinates were specified in PLAYMOL to build the initial configurations for simulating the liquid phases.

@@

\subsection{Simulation strategies of ionic liquids}
\label{sec:prel_results}
In this section, we present results of preliminary simulations using the  alternatives to standard approaches of modeling the ions as rigid bodies and employing the DSF method~\cite{Fennell2006} For that, we compare the simulated properties of density and radial distribution functions to that obtained when considering the ions as flexible molecules and the Ewald method for the Coulombic interactions.
